### 2019-01-16 submission/manuscript.Rmd
Pat's edits

# Assessing the Differences in 16S rRNA Gene Sequencing Due to High Fidelity DNA Polymerase

# Influence of DNA Polymerase on PCR bias, chimera formation, and errors when sequencing 16S rRNA genes from complex communities

\begin{center}
\vspace{25mm}

@@

@@

16S rRNA gene sequencing is a powerful and widely used tool for surveying the structure of microbial communities [@Gilbert2018; @HMP2012; @Schloss2016]. This approach has exploded in popularity with the advent of high throughput sequencing where it is possible to characterize numerous samples with thousands of sequences per sample. Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples [@Luo2016; @Bassis2017; @Gorzelak2015; @Dominianni2014; @BautistadelosSantos2016], storage conditions [@Sinha2015; @Amir2017b; @Lauber2010; @Luo2016; @Bassis2017; @Gorzelak2015; @Song2016], extraction methods [@Costea2017], amplification conditions [@BautistadelosSantos2016; @Haas2011; @Sinha2017], sequencing method [@Meisel2016; @Caporaso2010; @Sinha2017], and analytical pipeline [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017]. The increased sampling depth that is now available relative to previous Sanger-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been long known to have a significant impact on the description of microbial communities is the choice of conditions for PCR amplification [@BautistadelosSantos2016; @Haas2011; @Sinha2017]. Factors such as the choice of primers have an obvious impact on which populations will be amplified [@Parada2015; @Kozich2013]. However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras [@Haas2011; @Wang1996; @Potapov2017; @Kebschull2015], misincorporation of nucleotides [@Potapov2017; @McInerney2014; @Cline1996], preferential amplification of populations leading to bias [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998], and accumulation of random amplification events leading to PCR drift [@Kebschull2015; @Kennedy2014; @Acinas2005; @Wagner1994]. Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs [@Haas2011; @Edgar2011]. Others have attempted to account for PCR bias using modeling approaches [@Edgar2017; @Brooks2015]; however, these have been developed for idealized situations. Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability [@Haas2011; @Potapov2017]. To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift [@Kennedy2014]. Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community [@Polz1998; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998]. Early investigations of the factors that lead to the formation of PCR artifacts focused on analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation. Although these studies were instrumental in forcing researchers to be cautious about the interpretation of their results, we have a poor understanding of how these factors affect the formation of PCR artifacts in more complex communities.

The influence that the choice of DNA polymerase has on the formation of PCR artifacts has not been well studied. There has been recent interest in how the choice of the hypervariable region and data analysis pipelines impact the sequencing error rate [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017]; however, these studies use the same DNA polymerase in the PCR step and implicitly assume that the rate of nucleotide misincorporation from PCR are significantly smaller than those from the sequencing phase. There has been more limited interest in the impact that DNA polymerase choice has on the formation of chimeras [@Potapov2017; @Gohl2016]. A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@Gohl2016]. The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@Gohl2016]. Yet this optimization was specific for the community they were analyzing (i.e. captive and semi-captive red-shanked doucs) and assumed that if the two polymerases generate the same community structure that the community structure was correct. In fact, the community structure generated by both methods was not free of artifacts, but has the same artifacts. A challenge in these types of experiments is having a priori knowledge of the true community representation. A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias [@Schloss2011]; however, mock communities have a limited phylogenetic diversity relative to natural communities. Natural communities, in contrast, have an unknown community composition making absolute measurements impossible. They can be used to validate results from mock communities and to understand the relative impacts of artifacts on the ability to differentiate biological and methodological sources of variation. Given the large number of DNA polymerases available to researchers it is unlikely that a specific recommendation is possible. Rather, the development of general best practices and understanding the impact of PCR artifacts on an analysis are desired.

This study investigated the impact of choice of high fidelity DNA polymerase and the number of rounds of amplification on the formation of PCR artifacts using a mock community and human stool samples. These PCR artifacts included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process. To characterize these factors we sequenced a mock community of 8 organisms with known sequences and community structure and human fecal samples with unknown sequences and community structures. We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform. This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol.
To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 15, 20, 25, 30, and 35 rounds of amplification (Table 1). Collectively, our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals.

@@

***Sequencing errors vary by the number of cycles and the DNA polymerase used in PCR.*** The presence of sequence errors can confound the ability to accurately classify 16S rRNA gene sequences and group sequences into operational taxonomic units (OTUs). More importantly, sequencing errors themselves can alter the representation of the community. Therefore, it is important to minimize the number of sequencing errors. Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community. We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [@Kozich2013; @Schloss2011; @Quince2011]. Regardless of the polymerase, the error rate increased with the number of rounds of amplification (Figure 1). Using 30 rounds of PCR is a common approach across diverse types of samples. Among the data generated using 30 rounds of PCR the Accuprime polymerase had the highest error rate (i.e. `r filter(error, polymerase=="ACC") %>% select(contig_error)`%) followed by the Platinum (i.e. `r filter(error, polymerase=="PL") %>% select(contig_error)`%), Phusion (i.e. `r filter(error, polymerase=="PHU") %>% select(contig_error)`%), Kappa (i.e. `r filter(error, polymerase=="K") %>% select(contig_error)`%), and Q5 (i.e. `r filter(error, polymerase=="Q5") %>% select(contig_error)`%) polymerases (Figure 1). When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e. `r filter(error, polymerase=="PL") %>% select(pc_error)`%) followed by the Accuprime (i.e. `r filter(error, polymerase=="ACC") %>% select(pc_error)`%), Q5 (i.e. `r filter(error, polymerase=="Q5") %>% select(pc_error)`%), Phusion (i.e. `r filter(error, polymerase=="PHU") %>% select(pc_error)`%), and Kappa (i.e. `r filter(error, polymerase=="K") %>% select(pc_error)`%) polymerases (Figure 1). Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.

@@

***The fraction of sequences identified as being chimeric varies by the number of cycles and the DNA polymerase used in PCR.*** Chimeric PCR products can significantly confound downstream analyses. Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors. We identified those contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence. As expected from previous studies, the number of chimeras increased with rounds of amplification (Figure 2A). Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used. After 30 rounds of PCR, the Platinum polymerase had the highest chimera rate (i.e. `r filter(chimera, polymerase=="PL") %>% select(pc_frac_chimera)`%) followed by the Q5 (i.e. `r filter(chimera, polymerase=="Q5") %>% select(pc_frac_chimera)`%), Phusion (i.e. `r filter(chimera, polymerase=="PHU") %>% select(pc_frac_chimera)`%), Kappa (i.e. `r filter(chimera, polymerase=="K") %>% select(pc_frac_chimera)`%), and Accuprime (i.e. `r filter(chimera, polymerase=="ACC") %>% select(pc_frac_chimera)`%) polymerases. We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity (Figure 2A). For all polymerases, the specificity was above `r min_specificity`% and showed a weak association with the number of cycles used. There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community. This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity. The sensitivity of UChime varied between `r sensitivity["min"]` and `r sensitivity["max"]`%. The generalizability of these results is limited because we used a single mock community with limited genetic diversity. Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric  (Figure 2B). These results followed those from the mock communities; additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used. Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.

@@

***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses. The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria. The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community. We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase (Figure 3). Interestingly, for some bacteria, their representation became less biased with additional rounds of PCR (e.g. *L. fermentum*), while others became more biased (e.g. *E. faecalis*), and others had little change (e.g. *B. subtilis*). The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias. Across the strains there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed. One of the bacteria represented in the mock community, *S. enterica*, had 6 identical copies of the V4 region and 1 copy that differed from those by one nucleotide. The dominant copy had a thymidine and the rare copy had a guanine. We used the sequence data to calculate the ratio of the dominant to rare variants from *S. enterica* (Figure S1). The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of `r others`; however, the ratio for the Kappa polymerase varied between `r kappa["min"]` and `r kappa["max"]` for 25 to 35 rounds of PCR. Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias and the previous observation regarding the guanine and cytosine relative abundance was not observed with our data.

@@

***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the distance between samples and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we used analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline (Figure 4). First, we removed the chimeras from the mock community data as described above and mapped the individual reads to the OTUs that the 16S rRNA gene fragments would cluster into if there were no sequencing errors. This gave us a community distribution that reflected the distribution following PCR without any artifacts. Alpha diversity metrics such as richness and Shannon and Inverse Simpson diversity metrics increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, where the richness remained constant but the diversity decreased (Figure 4A). These data suggest that PCR had the effect of making the community distribution more even than it was originally. Next, we used the true sequence diversity, but removed chimeras based on mapping reads to all possible chimeras between reference sequences, and clustered the reads to OTUs. The same trends were observed as with perfect sequence data except that the richness and diversity metrics trended higher (Figure 4A). Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras. Again, we observed the same trends as with the other analyses except the richness and diversity metrics trended higher (Figure 4A). These comparisons demonstrated that although the bias at the sequence level was subtle, PCR does have a bias that is exacerbated by errors and chimeras. When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of `r mean_diff` units (Figure 4B). The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: `r filter(pl, col_round=="30x") %>% select(distances)`; 25 vs 35 rounds: `r filter(pl, col_round=="35x") %>% select(distances)`). Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR (Figure 4C). This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=`r filter(mock_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) was a larger factor than the choice of polymerase (R^2^=`r filter(mock_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001). These results demonstrate that subtle differences in relative abundances can have an impact on overall community structure. This variation underscores the importance of only comparing sequence data that have been generated using the same PCR conditions.

@@

***There is little evidence of a relationship between polymerase or number of rounds of amplification on PCR drift.*** There have been concerns that the same template DNA subjected to the same PCR conditions could result in different representations of communities because of random drift. To test this, we determined the average Bray-Curtis distance between replicate reactions using the same polymerase and number of rounds of amplification (Figure 6). Using the mock community data there were no obvious trends. The average Bray-Curtis distance within a set of conditions varied by `r mock_drift["min"]` to `r mock_drift["max"]` units. Although we did not obtain replicates of each of the stool samples, the intra-sample variation for each set of conditions was consistent and varied between `r stool_drift["min"]` and `r stool_drift["max"]` units. These data suggest that amplicon sequencing is robust to random variation in amplification and that any differences are likely to be smaller than what is considered biologically relevant.

@@

Our results suggest that the number of rounds of PCR and to a lesser degree the choice of DNA polymerase impact the analysis of 16S rRNA gene sequence data from bacterial communities. Although it was not possible to make direct connections between PCR conditions and specific sources of bias, we were able to identify general recommendations that reduce the amount of error, chimera formation, and bias. Researchers should strive to minimize the number of rounds of PCR and should use a high fidelity polymerase. Although specific PCR conditions impact the precise interpretation of the data, the effects were consistent and were smaller than the biological differences between individuals. Based on these observations, within a study, amplicons must be generated by consistent protocols to yield meaningful comparisons. When comparing across studies, values like richness, diversity, and relative abundances must be made in relative and not absolute terms.

@@

***Statistical analysis.***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2018]. Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2017]. The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`) [@Vegan2018].

@@

@@

@@

>>>>

\vspace{25mm}
\begin{center}

\textbf{\huge Quantifying the impact of DNA polymerase and number of rounds of amplification in PCR on 16S rRNA gene sequence data}

\vspace{50mm}

\textbf{Running title:} Quantifying the effects of PCR

\vspace{25mm}

@@

PCR amplification of 16S rRNA genes is a critical, yet under appreciated step in the generation of sequence data to describe the taxonomic composition of microbial communities. Numerous factors in the design of PCR can impact the sequencing error rate, the abundance of chimeric sequences, and the degree to which the fragments in the product represent their abundance in the original sample. The choice of PCR cycling conditions are known to impact the outcome of PCR and downstream inferences. We compared the performance of high fidelity polymerases and varying number of rounds of amplification when amplifying a mock community with known sequences and community structure and human stool samples. Although it was impossible to derive specific recommendations, we did observe general trends. Namely, using a polymerase with the highest possible fidelity and minimizing the number of rounds of PCR reduced the sequencing error rate, fraction of chimeric sequences, and bias. Evidence of bias at the sequence level was subtle and could not be ascribed to the fragments' fraction of bases that were guanines or cytosines. When analyzing mock community data, the amount that the community deviated from the expected composition increased with rounds of PCR. The effect was inconsistent for human stool samples. Overall the results underscore the difficulty of comparing sequence data that are generated by different PCR protocols. However, the results indicate that the variation in human stool samples is generally larger than that introduced by the choice of polymerase or number of rounds of PCR.

### Importance

A steep decline in sequencing costs drove an explosion in studies characterizing microbial communities from diverse environments. Although a significant amount of effort has gone into understanding the error profiles of DNA sequencers, little has been done to understand the downstream effects of the PCR amplification protocol. We quantified the effects of the choice of polymerase and number of PCR cycles on the quality of downstream data. We found that these choices can have a profound impact on the way that a microbial community is represented in the sequence data. The effects are relatively small compared to the variation in human stool samples, however, care should be taken to use polymerases with the highest possible fidelity and to minimize the number of rounds of PCR. These results underscore the difficulty of directly comparing the sequence data generated using different PCR protocols or interpreting results in an absolute sense.

@@

16S rRNA gene sequencing is a powerful and widely used tool for surveying the structure of microbial communities [@Gilbert2018; @HMP2012; @Schloss2016]. This approach has exploded in popularity with the advent of high throughput sequencing where it is possible to characterize numerous samples with thousands of sequences per sample. Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples [@Luo2016; @Bassis2017; @Gorzelak2015; @Dominianni2014; @BautistadelosSantos2016], storage conditions [@Sinha2015; @Amir2017b; @Lauber2010; @Luo2016; @Bassis2017; @Gorzelak2015; @Song2016], extraction methods [@Costea2017], amplification conditions [@BautistadelosSantos2016; @Haas2011; @Sinha2017], sequencing method [@Meisel2016; @Caporaso2010; @Sinha2017], and analytical pipeline [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017]. The increased sampling depth that is now available relative to previous Sanger sequencing-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been long known to have a significant impact on the description of microbial communities is the choice of conditions for PCR amplification [@BautistadelosSantos2016; @Haas2011; @Sinha2017]. Factors such as the choice of primers have an obvious impact on which populations will be amplified [@Parada2015; @Kozich2013]. However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras [@Haas2011; @Wang1996; @Potapov2017; @Kebschull2015], misincorporation of nucleotides [@Potapov2017; @McInerney2014; @Cline1996], preferential amplification of populations leading to bias [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998], and accumulation of random amplification events leading to PCR drift [@Kebschull2015; @Kennedy2014; @Acinas2005; @Wagner1994]. Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs [@Haas2011; @Edgar2011]. Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability [@Haas2011; @Potapov2017]. Others have attempted to account for PCR bias using modeling approaches [@Edgar2017; @Brooks2015]; however, these have been developed for idealized situations. To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift [@Kennedy2014]. Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community [@Polz1998; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998]. Early investigations of the factors that lead to the formation of PCR artifacts focused on analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation. Although these studies were instrumental in forcing researchers to be cautious about the interpretation of their results, we have a poor understanding of how these factors affect the formation of PCR artifacts in more complex communities.

The influence that the choice of DNA polymerase has on the formation of PCR artifacts has not been well studied. There has been recent interest in how the choice of the hypervariable region and data analysis pipelines impact the sequencing error rate [@Kozich2013; @Schloss2011; @Bokulich2012; @Sinha2017]; however, these studies use the same DNA polymerase in the PCR step and implicitly assume that the rate of nucleotide misincorporation from PCR are significantly smaller than those from the sequencing phase. There has been more limited interest in the impact that DNA polymerase choice has on the formation of chimeras [@Potapov2017; @Gohl2016]. A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@Gohl2016]. The authors of the study reduced the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@Gohl2016]. Yet this optimization was specific for the community they were analyzing (i.e. captive and semi-captive red-shanked doucs) and assumed that if the two polymerases generate the same community structure that the community structure was correct. In fact, the community structure generated by both methods was not free of artifacts, but likely had the same artifacts. A challenge in these types of experiments is having *a priori* knowledge of the true community representation. A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias [@Schloss2011]; however, mock communities have a limited phylogenetic diversity relative to natural communities. Natural communities, in contrast, have an unknown community composition making absolute measurements impossible. They can be used to validate results from mock communities and to understand the relative impacts of artifacts on the ability to differentiate biological and methodological sources of variation. Given the large number of DNA polymerases available to researchers, it is unlikely that a specific recommendation is possible. Rather, the development of general best practices and understanding the impact of PCR artifacts on an analysis are needed.

This study investigated the impact of choice of high fidelity DNA polymerase and the number of rounds of amplification on the formation of PCR artifacts using a mock community and human stool samples. These PCR artifacts included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process. To characterize these factors we sequenced a mock community of 8 organisms with known sequences and community structure and human fecal samples with unknown sequences and community structures. We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform. This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol [@Kozich]. To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 20, 25, 30, and 35 rounds of amplification. Collectively, our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals.

@@

***Sequencing errors vary by the number of cycles and the DNA polymerase used in PCR.*** The presence of sequence errors can confound the ability to accurately classify 16S rRNA gene sequences and group sequences into operational taxonomic units (OTUs). More importantly, sequencing errors themselves can alter the representation of the community. Therefore, it is important to minimize the number of sequencing errors. Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community. We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [@Kozich2013; @Schloss2011; @Quince2011]. Regardless of the polymerase, the error rate increased with the number of rounds of amplification (Figure 1). Using 30 rounds of PCR is a common approach across diverse types of samples. Among the data generated using 30 rounds of PCR the Accuprime polymerase had the highest error rate (i.e. `r filter(error, polymerase=="ACC") %>% select(contig_error)`%) followed by the Platinum (i.e. `r filter(error, polymerase=="PL") %>% select(contig_error)`%), Phusion (i.e. `r filter(error, polymerase=="PHU") %>% select(contig_error)`%), Kappa (i.e. `r filter(error, polymerase=="K") %>% select(contig_error)`%), and Q5 (i.e. `r filter(error, polymerase=="Q5") %>% select(contig_error)`%) polymerases (Figure 1). When we applied a pre-clustering denoising step, which merged the counts of reads within 2 nt of a more abundant sequence [@Schloss2011], the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e. `r filter(error, polymerase=="PL") %>% select(pc_error)`%) followed by the Accuprime (i.e. `r filter(error, polymerase=="ACC") %>% select(pc_error)`%), Q5 (i.e. `r filter(error, polymerase=="Q5") %>% select(pc_error)`%), Phusion (i.e. `r filter(error, polymerase=="PHU") %>% select(pc_error)`%), and Kappa (i.e. `r filter(error, polymerase=="K") %>% select(pc_error)`%) polymerases (Figure 1). Although specific recommendations are difficult to make because the phylogenetic diversity of the initial DNA template is likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.

@@

***The fraction of sequences identified as being chimeric varies by the number of cycles and the DNA polymerase used in PCR.*** Chimeric PCR products can significantly confound downstream analyses. Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors [@Edgar2011; @Haas2011]. Because the true sequences of the organisms in the mock community were known, we could generate all possible chimeras between pairs of V4 16S rRNA gene fragments and use these possible chimeric sequences to screen the sequences generated under the different PCR conditions for chimeras. The number of chimeras increased with rounds of amplification (Figure 2A). Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used. After 30 rounds of PCR, the Platinum polymerase had the highest chimera rate (i.e. `r filter(chimera, polymerase=="PL") %>% select(pc_frac_chimera)`%) followed by the Q5 (i.e. `r filter(chimera, polymerase=="Q5") %>% select(pc_frac_chimera)`%), Phusion (i.e. `r filter(chimera, polymerase=="PHU") %>% select(pc_frac_chimera)`%), Kappa (i.e. `r filter(chimera, polymerase=="K") %>% select(pc_frac_chimera)`%), and Accuprime (i.e. `r filter(chimera, polymerase=="ACC") %>% select(pc_frac_chimera)`%) polymerases. Because our chimera screening procedure can only be applied to mock communities, we used the UCHIME algorithm to model the chimera screening approach that is used in most sequence curation pipelines. By comparing the output of UCHIME to our reference-based approach, we were able to calculate the UCHIME's sensitivity and specificity (Figure 2A). The specificity for all polymerases was above `r min_specificity`% and showed a weak association with the number of cycles used. There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UCHIME to detect the chimeras from the mock community. This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity. The sensitivity of UCHIME varied between `r sensitivity["min"]` and `r sensitivity["max"]`%. The generalizability of these results is limited because we used a single mock community with limited genetic diversity. Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UCHIME identified as being chimeric  (Figure 2B). These results followed those from the mock communities: additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used. Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.

@@

***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to amplify 16S rRNA gene fragments there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses [@Kebschull2015; @Acinas2005; @Polz1998; @Brooks2015; @Suzuki1996; @Chandler1997; @Wagner1994; @Hansen1998]. The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria because each bacterium had a different genome size and varied in the number of operons in its genome. The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community. We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase (Figure 3). Interestingly, for some bacteria, their representation became less biased with additional rounds of PCR (e.g. *L. fermentum*), while others became more biased (e.g. *E. faecalis*), and others had little change (e.g. *B. subtilis*). Contrary to prior reports [@Polz1998], The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias. Across the strains there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed. One of the bacteria represented in the mock community, *S. enterica*, had 6 identical copies of the V4 region and 1 copy that differed from those by one nucleotide. The dominant copy had a thymidine and the rare copy had a guanine. We used the sequence data to calculate the ratio of the dominant to rare variants from *S. enterica* (Figure S1). The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of `r others`; however, the ratio for the Kappa polymerase varied between `r kappa["min"]` and `r kappa["max"]` for 25 to 35 rounds of PCR. Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias.

@@

***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the variation in bias between polymerases and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline (Figure 4). First, we removed the chimeras from the mock community data as described above and mapped the individual reads to the OTUs that the 16S rRNA gene fragments would cluster into if there were no sequencing errors. This gave us a community distribution that reflected the distribution following PCR without any artifacts. Although the richness did not change, the Shannon diversity increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, for which the diversity decreased (Figure 4A). These data suggest that PCR had the effect of making the community distribution more even than it was originally, except for the data generated using the Kappa polymerase where the evenness decreased. Next, we used the observed sequence errors, but removed chimeras using our reference-based approach, and clustered the reads to OTUs. The richness and diversity metrics trended higher with higher error rates and number of rounds of PCR (Figure 4A). Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras. Again, the richness and diversity metrics trended higher with higher error rates and number of rounds of PCR (Figure 4A). These comparisons demonstrated that although the bias at the sequence level was subtle, PCR introduces bias that is exacerbated by errors and chimeras when sequences are clustered into OTUs. When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of `r mean_diff` units (Figure 4B). The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: `r filter(pl, col_round=="30x") %>% select(distances)`; 25 vs 35 rounds: `r filter(pl, col_round=="35x") %>% select(distances)`). Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR (Figure 4C). This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=`r filter(mock_vegan, effects=="rounds") %>% select(r_sq)`, P<0.001) was a larger factor than the choice of polymerase (R^2^=`r filter(mock_vegan, effects=="polymerase") %>% select(r_sq)`, P<0.001). These results demonstrate that subtle differences in relative abundances can have an impact on overall community structure. This variation underscores the importance of only comparing sequence data that have been generated using the same PCR conditions.

@@

***There is little evidence of a relationship between polymerase or number of rounds of amplification on PCR drift.*** There have been concerns that the same template DNA subjected to the same PCR conditions could result in different representations of communities because of random drift over the course of PCR. To test this, we determined the average Bray-Curtis distance between replicate reactions using the same polymerase and number of rounds of amplification (Figure 6). Using the mock community data there were no obvious trends. The average Bray-Curtis distance within a set of conditions varied by `r mock_drift["min"]` to `r mock_drift["max"]` units. Although we did not generate technical replicates of each of the stool samples, the inter-sample variation for each set of conditions was consistent and varied between `r stool_drift["min"]` and `r stool_drift["max"]` units. These data suggest that amplicon sequencing is robust to random variation in amplification and that any differences are likely to be smaller than what is considered biologically relevant.

@@

Our results suggest that the number of rounds of PCR and to a lesser degree the choice of DNA polymerase impact the analysis of 16S rRNA gene sequence data from bacterial communities. Although it was not possible to make direct connections between PCR conditions and specific sources of bias, we were able to identify general recommendations that reduce the amount of error, chimera formation, and bias. Researchers should strive to minimize the number of rounds of PCR and should use a high fidelity polymerase. Although specific PCR conditions impact the precise interpretation of the data, the effects were consistent and were smaller than the biological differences between individuals. Based on these observations, amplicons must be generated by consistent protocols to yield meaningful comparisons. When comparing across studies, values like richness, diversity, and relative abundances must be made in relative and not absolute terms.

@@

***Statistical analysis.***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2018]. Data transformation and graphing were completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`). The distance matrix data was analyzed using the adonis function within the vegan package (v `r paste(packageVersion("vegan")[1])`).

@@

\includegraphics{../results/figures/mock_error.pdf}

@@

\includegraphics[width=1.0\columnwidth]{../results/figures/chimera_plots.pdf}

@@

\includegraphics[width=1.0\columnwidth]{../results/figures/salmonella_bias.pdf}
